Title Goes Here

### 2024-01-05 analysis/paper/paper.qmd
- First (first) draft of the article.

---
title: "Title Goes Here"
author:
  - Jane Doe:
      correspondence: "yes"
      email: janedoe@fosg.org
      orcid: 0000-0003-1689-0557
      institute:
        - fosg
        - fop
  - John Q. Doe:
      institute: fosg
      orcid: 0000-0003-1689-0558
  - Peder Ås:
      institute: fosg
      orcid: 0000-0003-1689-0559
  - Juan Pérez:
      orcid: 0000-0003-1689-0551
      institute:
        - name: Acme Corporation
  - Max Mustermann:
      orcid: 0000-0003-1689-0552
institute:
  - fosg:
      name: Formatting Open Science Group
      address: 23 Science Street, Eureka, Mississippi, USA
  - fop: Federation of Planets
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

@@

```{r packages, echo=FALSE}
# Call packages
library(readr)
library(dplyr)
library(stringr)
library(tidyr)
library(ggplot2)
library(scales)
library(janitor)

@@

Here is a citation [@Marwick2017]

# Background

# Methods

The samples chosen for this study consist of the chert and chalcedony lithic artifacts with individual IDs (excluding buckets) from the Upper Paleolithic levels of the Terrace and Shelter areas of Vale Boi. The sample from the Terrace area (VB-T) consists of 3653 from levels 8 to 4, while the sample from the Shelter area (VBS) consists of 1652 from levels C to A. These two areas were chosen since both have detailed spatial information (as mentioned in section *IN*), several radiocarbon dates through different levels, and previous studies providing lithic technology, faunal and geoarchaeological data.

## Geo-archaeological surveys

The geological data and comparative collection from local and regional cherts used in this study were obtained through previous prospections and petrographic works (\*\*), located at the Lusolit lithotheque (hosted at ICArEHB, University of Algarve). These samples also included cherts collected from previous prospections from other areas of Portugal (such as central Portugal). Hand samples were used throughout the macroscopic characterization for comparison with the archaeological materials, and the archaeological thin sections were compared with the previously studied geological chert thin sections. The lithotheque from the Unit of Geoarchaeology and Archaeometry Applied to Historic Artistic and Monumental Heritage from the University of Cadiz was also visited, to understand the macroscopic variability from the Cadiz local and regional cherts, to discern between local and regional chert types and possible long-distance cherts.

## Macroscopic characterization

We applied a macroscopic analysis to both Terrace and Shelter samples, which was divided into two steps. The first step included a preliminary characterization using a hand lens of 10x magnification. This step allowed us to create groups and subgroups based on similar macroscopic characteristics which have been used in previous studies to describe and discern between different types of chert; these include characteristics such as color, translucency, and feel; a data dictionary for the recorded variables can be found in the Supplementary Materials (\*\*).

In the second step, we used a stereomicroscope () to observe each sample in further detail, which allowed us to better characterize the artifacts (especially regarding inclusions and alterations), and the previously established groups. The data dictionary used for the individual samples can also be found in Supplementary Materials (\*\*). Given the inherent macroscopic variability of the geological samples and archaeological specimens, as well as the number of samples, the database used for the individual analysis is simplified and focused on collecting data related to weight, cortex, and alterations.

## Petrographic characterization

To better characterize the different chert groups identified in the macroscopic analysis, we analyzed ++ thin sections. The samples chosen for this analysis include at least one sample from each group present throughout the stratigraphy of the Terrace area, from different archaeological levels. For groups which showed macroscopic variability, several thin sections were analysed to better characterize these cherts. @tbl-TS-table Petrographic thin sections were analyzed by polarized light microscopy () to determine the mineral composition, textural characteristics, allochems, and bioclasts, as well as any alterations. A total of 20 thin sections were prepared at Thin Section Lab (Toul, France) and analyzed at the ICArEHB laboratory. The archaeological samples chosen for the petrographic study are from the Terrace area, from several levels, and focused on characterizing the macroscopic variability, reflected on the several groups and subgroups (table \*\*).

```{r}
#| label: tbl-TS-table
#| tbl-cap: Thin section list.
#| echo: false

TS_short_list <- read_csv(here::here('analysis/data/raw_data/TS_short_list.csv'))

knitr::kable(TS_short_list)

@@

@@

T_dataset <- T_dataset %>% 
  #filter(Level %in% c("4", "4C", "4D", "4E", "5", "6", "7"))
  filter(Level !="4B" & Level !="7C" & Level!="8") # Filtering out levels with barely any artifacts

cols_to_concat_spit <- c("Level", "Spit")
RM_by_spit <- T_dataset %>%
  filter(!Spit %in% c("limp_corte", "profile", " ", NA)) %>%
  select(ID, Z, Level, Spit, RMTYPE) %>%
  unite(col='depth', cols_to_concat_spit, sep=".", remove=FALSE) %>% 
  dplyr::group_by(depth) %>%
  dplyr::count(RMTYPE) %>%
  mutate(freq = n / sum(n)*100)

@@

# Shelter DB preparation
S_RM_data <- read_csv(here::here('analysis/data/raw_data/MACRO_IND_SHELTER.csv')) # Shelter RM analysis
S_field_data <- read_csv(here::here('analysis/data/raw_data/VB_Shelter_xyz.csv')) # Shelter coordinates

@@

S_field_data <- S_field_data %>%
  unite(col='ID', cols_to_concat_shelter, sep="-", remove=FALSE) %>% # Unite YEAR with Id s«using cols_to_concat
  select(ID, X, Y, Z) %>% # Selecting important variables
  distinct(ID, .keep_all = TRUE)
S_dataset <- left_join(S_RM_data, S_field_data, by = "ID")

@@

The three main used raw materials at the archaeological site of Vale Boi are quartz, greywacke, and chert. Using the field database with all individually coordinated lithic artifacts, we compared the incidence of chert and chalcedony in opposition to the remaining lithics of other raw materials. Since some lithic assemblages of the Terrace area are currently under study, the data regarding the percentage of raw materials in the “Other” category was not available for all archaeological levels. @fig-rm-comparison shows that for the selected levels, and using single coordinated artifacts (excluding buckets), chert represents only \~20-15% of the lithic assemblage. Based on previous lithic studies in the terrace area, the remaining 80% of other raw materials are mostly composed of quartz and greywacke. Although no systematic raw material research has been applied to the latter raw materials, they have been interpreted as local and readily available in the proximity of the site. Frequently, the quartz and greywacke assemblage includes large amounts of shattered and/or unknapped chunks or slabs. Previous studies from the terrace area have also shown that when removing shatter and chunks, the percentages of chert become much more representative when compared to greywacke and quartz. Based on this, despite the low amounts of individually plotted chert artifacts in the field database, chert continues to be one of the main raw materials to produce lithic stone tools, with complete knapping sequences and formal toolkits.

indet_table <- phase_dataset %>% 
  mutate(Group = if_else(RMTYPE=="INDET", "Indeterminate", "Grouped")) %>% 
  tabyl(Group,Level) %>% 
  adorn_totals(c("row", "col")) %>%
  adorn_percentages("col") %>%
  adorn_pct_formatting(2) %>%
  adorn_ns("front")

As previously mentioned, whenever too altered to confidently attribute a group, the artifacts were instead classified as indeterminable (INDET). As seen in @tbl-indet-freq, level 4 shows the higher percentages of indeterminate artifacts, corresponding to 40% of the total sample (n=1027). This is due to the frequent alterations of the sample, especially related with patinas and alteration rinds, pits and fire/heat alterations. For all other levels, indeterminate represents in mean \~28% of the chert artifacts.

@@

#T_short_table <- T_short %>% 
#   tabyl(Rawmaterial, Level) %>%
# adorn_totals(c("row", "col")) %>%
#  adorn_percentages("col") %>%
#  adorn_pct_formatting(2) %>%
#  adorn_ns("front")

@@

Through the macroscopic and petrographic analysis we identified 24 types of chert and three types of chalcedony. Of these, one type of chalcedony and 9 types of chert are composed of several samples (\>10) and present across both areas (Terrace and Shelter) and throughout the Upper Paleolithic occupations of the Terrace area. Thin sections were produced for the lithics of the latter groups. One type had several samples (\>10; Type GB) but was only identified in the Shelter area, while 14 varieties of chert and 2 chalcedony are composed of one to three samples and restricted to specific levels. The latter varieties were collapsed into a category named TL (Trace lithotypes).

We present here a short description of each identified macroscopic and petrographic type. A detailed macroscopic and petrographic description and figures for each type and sub-type of chert and chalcedony can be found in the SOM (++). The petrographic analysis PDF files for all thin sections studied in this paper (including macroscopic and microscopic figures) can also be found in our GitHub page (++) as well as the OSF repository (++).

## Type description

The chalcedony is a massive, fibrous chalcedony mudstone. It is characterized by a colorless and translucent appearance, often partially or completely white patinated. It has good to medium quality, due to frequent large crystallizations, irregularities or fractures which impact the knapping. In thin section, the chalcedony is composed by fibrous chalcedony, without allochems or bioclasts, and no porosity (\<1%). Radiating spherulitic textures are very frequent.

The first chert groups identified at Vale Boi are massive micro-cryptocrystalline quartzes from a marine environment. Petrographically, these cherts are homogeneous, except for a small amount of artifacts with a banded variety within type 12. These cherts are characterized by wackestone textures, with accessory amounts of macrocrystalline quartz and fibrous chalcedony, frequently found replacing fossils and occasionally dolomite. Group 12 shows higher percentages of micrite/sparite. In all samples, fossils range from common to very frequent, albeit frequently unidentifiable. When identifiable in thin-section, they are sponge spicules, with the rare occurrence of possible foraminifera and ostracod in a sample from type 12. Porosity varies between 10% and 1%, of vug type. Macroscopically however, these groups show variability, which resulted in their separation in 4 different macroscopic groups; this separation was maintained for its possible usefulness in identifying preferred types of nodules or macroscopic varieties through time and for future correlation with technological data.

Group 2 is characterized by single-color and multiple-color samples, ranging between yellow, grayish brown, pale red, and white, and with variable patterns such as splotched, broad mottling, speckling, and laminations. They are opaque, with sub-translucent areas mainly focused in the white areas of the sample. Whenever present, the cortex can be rounded or irregular, varying between thin to thick. Inclusions include black speckles and red speckles, which are probably organics and iron oxides. Bioclasts are frequently unidentifiable, but also include monaxon sponge spicules, bivalve shells replaced by quartz or chalcedony, and rarely triaxon sponge spicules. In the shelter area, the variety of identified fossils also includes possible uncommon ostracods, a possible small foraminifer, and echinoderm spines.

Group 15 is characterized by light gray and reddish gray colors and varies between opaque to sub-translucent. It is characterized by the presence of splotches (often coarser and sub-translucent), speckling, and fine lamination. Whenever present, the cortex is white, rounded, and of medium thickness. Allochems such as iron oxides or opaques are also visible macroscopically as black speckling or rare red speckling. The bioclasts are frequently unidentifiable, due to poor preservation and are replaced by microcrystalline quartz or fibrous chalcedony, except for uncommon sponge spicules.

Type 7 shows a heterogeneous structure with two types of fabrics. The interior of the samples and the nodules is gray, and sub-translucent with a single mica mineral, flecks, and speckles. The outer edge and within the grey fabric as splotches or broad mottling are white, opaque, and with yellow speckles. The cortex is white, with a sharp transition and with a medium to thick width. The morphology of the cortex in fully cortical samples is originally available as irregular (slightly oval) nodules within the parent rock. Whenever identifiable, bioclasts are limited to sponge spicules.

>>>>

---
title: "Title Goes Here"
author:
  - Jane Doe:
      correspondence: "yes"
      email: janedoe@fosg.org
      orcid: 0000-0003-1689-0557
      institute:
        - fosg
        - fop
  - John Q. Doe:
      institute: fosg
      orcid: 0000-0003-1689-0558
  - Peder Ås:
      institute: fosg
      orcid: 0000-0003-1689-0559
  - Juan Pérez:
      orcid: 0000-0003-1689-0551
      institute:
        - name: Acme Corporation
  - Max Mustermann:
      orcid: 0000-0003-1689-0552
institute:
  - fosg:
      name: Formatting Open Science Group
      address: 23 Science Street, Eureka, Mississippi, USA
  - fop: Federation of Planets
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
number-sections: true
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
---

@@

```{r packages, echo=FALSE}
# Call packages
library(readr)
library(dplyr)
library(stringr)
library(tidyr)
library(ggplot2)
library(scales)
library(janitor)
library(ggpubr)
library(knitr)

@@

Knappable raw materials are a key element used to understand the lifeways of past hunter-gatherers throughout time, as lithics are ubiquitous throughout prehistory. More than mere stones with which stone tools are produced, these resources are intimately connected to the group’s technological, social, and cultural organization, possibly with an impact on the group's survivability [@oestmoFormalModelingApproach2017]. Several key topics to understand hunter-gatherer behavior and organization have been explored through raw material analysis, such as modalities of procurement and mobility strategies [@ambrose_social_1990; @binford_organization_1979; @binford_righteous_1985; @gould_lithic_1985; @gould_lithic_1985; @kuhn_unpacking_1991; @mccall_behavioral_2007], occupation types and their duration [@kuhn_upper_2004; @surovell_toward_2009]) to the establishment and dimension of social networks [@whallon_social_2006] and exchange between groups or individuals [@wurzHowiesonsPoortBacked1999; @gamblePalaeolithicSocietiesEurope1999]. This has frequently been done through the systematic characterization of raw materials (both geologically available and archaeological), identification of corresponding sources for a clear differentiation between local and long-distance raw materials, and the study of their proportions within a site, often throughout different occupations and chronologies. These approaches and concepts have been extensively applied to the study of lithic assemblages from prehistoric archaeological sites in the Iberian Peninsula (++). Frequently coupled with technological analyses, the study of raw materials, especially chert, through a wide range of methods, such as macroscopic, petrographic, and geochemical, have recurrently aided in answering key questions to Iberian archaeology, such as the disappearance of Neanderthals and consequent replacement by Modern Humans [@aubry_far_2022]. Raw material studies have also contributed to understanding how different groups during different chronological periods organized their technology, managed time, and risk (through the embedded/direct procurement approach), and chose resources [@matias2016; @matias_o_2012; @pereira_raw_2021; @pereiraLithicEconomyTerritory2016]. Furthermore, these studies frequently highlight long-distance procurement even if in low proportions at times. In fact, raw material studies and the distinction between local/regional and exotic have also been successfully applied to identify long-distance contacts (\>150 km) during the Upper Paleolithic in the Côa Valley which has been interpreted as the result of reciprocal exchange [@aubry_raw_2004-1; @aubry_upper_2016]. Although commonly known and identified during late prehistory [@nocete_circulation_2005; @rodriguezPrehistoricFlintExploitations2011; @costa_lithic_2022], the identification of these contacts has expanded our understanding of often large-scale social networks and contacts during the Upper Paleolithic in central Portugal, to paint a picture previously outlined of a wide interconnected territory. However, unlike other regions of the Iberian Peninsula such as central Portugal, and despite the potential of raw material studies to provide key data on the lifeways and organization of Paleolithic hunter-gatherers through time, as well as the mobility range and existence of social networks, the behaviors of management and procurement of raw materials are still in a preliminary phase in southern Portugal, especially in the Algarve region. Although much is known about the technological organization of Paleolithic hunter-gatherer groups in southwestern Portugal, especially due to the archaeological site of Vale Boi (see ++), raw material studies have been scarcer. Previous studies applied macroscopic methods to characterize the cherts from Vale Boi. These preliminary results have shown the presence of a large component of local cherts from the Jurassic limestone formations in conjunction with a less abundant and variable allochthonous chert component throughout the several Upper Paleolithic occupations of the site [@bicho_ecodynamics_2013; @pereira_territory_2016-1; @verissimo_jazidas_2004]. @pereira_territory_2016-1 suggest that the exogenous group may have been the result of embedded procurement within a 16 km radius, while the allochthonous group may have been the result of transportation during trips to other areas of the Algarve, Alentejo, or even south Spain. The presence of chert from the Cretaceous formations of central Portugal has also been suggested [@bicho_ecodynamics_2013]. Furthermore, previous studies aiming to characterize the local and regional chert sources in the Algarve region have established a comparative macroscopic and petrographic database [@belmiro2023; @pereira_presenting_2016-1], a key element to raw material studies, especially to distinguish between local and non-local raw materials and understanding changes in the archaeological record [@pop_simulating_2015; @surovell_toward_2009]. These preliminary results and the existence of regional reference collections show the potential for a systematic raw material study at the site, using different methods and approaches, applied to the lithic assemblages throughout the several Upper Paleolithic occupations, to understand how these hunter-gatherer groups behaved through time: how they explored their landscape and lithic resources, where they moved across a territory nestled between water and land, and with whom did they establish social networks. To answer these questions, we aim to characterize the raw material procurement and mobility patterns of hunter-gatherer groups throughout the Upper Paleolithic of southern Portugal. By studying the chalcedony and chert artifacts of the UP levels of Vale Boi we aim to characterize these raw materials macroscopically and petrographically, to identify different chert groups and their respective sources.

## Site description

Vale Boi is located on the western coast of the Algarve region (south Portugal), in a small valley that connects south to the Atlantic current coast at around 2 km and is bordered by a limestone hill. This hill is characterized by limestone exposures that form rocks shelters facing west and southwest [@cascalheira_tecnologia_2010]. The site extends for more than 10 000 sq m through the slope of the valley, through which three main occupation areas have been identified: Slope, Shelter and Terrace. The excavation methodology included the recording of all artifacts with dimensions superior to 2 cm or complete small artifacts (i.e., bladelets) with a total station, and was applied to all main areas. Several occupations have been identified in both the Terrace and the Shelter area. The Terrace includes occupations from the Holocene to the Upper Paleolithic, resulting in the identification of eight main lithostratigraphic units. The UP occupations start with the Gravettian (levels 8-6), followed by Proto-solutrean (bottom of 5 and 5/4E, sensu Belmiro et al., 2021), and Solutrean occupations (4D-4 and 3B). The Shelter area shows four main lithostratigraphic units, with Magdalenian, Solutrean and Gravettian occupations. From these, the Solutrean levels show the largest occupation, with three, well-preserved archaeological horizons [@cascalheira_vale_2013].

## Geological context

The Algarve region is a complex territory, characterized by several geomorphic sub-regions and geological units, which provide different lithological resources used throughout prehistory. There are two main geological units, the South Portuguese Zone (SPZ) in the north sector of Algarve and the Algarve basin to the south. The SPZ is characterized by schists, greywacke and quartzites, while the Algarve basin is characterized by limestones, dolomites, and marls. It is within these limestones and dolomitized limestones from the Triassic to the Upper Jurassic where chalcedony and chert nodules and beds can be found. Chalcedony is located in a 15 km radius from Vale Boi, in the marl-carbonated Triassic formations as lenses or nodules and was formed through hydrothermal processes [@bicho_ecodynamics_2013]. The cherts can be found in different Jurassic formations throughout the Algarve, but are most frequently known for the Lower Jurassic beds and small nodules in the limestone cliffs at ++ km distance from the site, which have been previously studied and characterized [@ribeiro_evolucao_2005; @belmiro2023]. Chert is also available in the less known Upper Jurassic limestone formation in western Algarve (++ km from the site), and in the center and eastern portion of the Algarve, in Middle and Upper Jurassic formations with chert nodules [@belmiro2023].

# Materials and methods

To characterize the siliceous raw materials in the Upper Paleolithic occupations of Vale Boi and identify their geological sources, we applied a macroscopic and petrographic approach to the study of the chalcedony and chert archaeological materials from the Terrace and Shelter area. We applied two different methods to obtain a more complete characterization of the materials and compensate for the inherent caveats of each methods [@luedtke_archaeologists_1992]. Macroscopic analysis, for example, shows several caveats related to the lack of quantitative variables [@bustillo_is_2009]. However, given the large sample size of chalcedony and chert lithics at the archaeological site, a macroscopic approach presents a non-destructive, easy and less costly of characterizing all the materials. As such, we apply a first step of macroscopic analysis, followed by a petrographic study to reduce the subjectivity of the analysis [@bustillo_is_2009] and complete the characterization with petrographic data. Similarly, the combination of these two methods was previously applied successfully in the characterization of the cherts from the Algarve region as the result of recent prospection works [@belmiro2023], and as such, a comprehensive macroscopic and petrographic comparative collection is available for the comparison with archaeological materials when using the same methods.

This reference collection is located at the Lusolit lithotheque (hosted at ICArEHB, University of Algarve), and aided in the characterization and source attribution of cherts. Hand samples were used throughout the macroscopic characterization for comparison with the archaeological materials, and the archaeological thin sections were compared with the previously studied geological chert thin sections. The geological samples at the lithotheque also included cherts collected from previous prospections from other areas of Portugal (such as central Portugal) and chalcedony, of which 4 thin sections were produced for this study (3 of chert from the Estremadura and ++ of chalcedony from the Algarve). The lithotheque from the Unit of Geoarchaeology and Archaeometry Applied to Historic Artistic and Monumental Heritage from the University of Cadiz (hereon UCA) was also visited, to understand the macroscopic variability of the Cadiz local and regional cherts, to aid in the source attribution of cherts not congruent with local sources and identify possible long-distance chert-use.

As aforementioned, the study started with a macroscopic analysis of archaeological materials. The samples chosen for this study consist of the chert and chalcedony lithic artifacts with individual IDs (excluding buckets) from the Upper Paleolithic levels of the Terrace and Shelter areas of Vale Boi.This means that the smaller fraction of artifacts (lithics smaller than 2 cm) was not considered, and thus we may be missing patterns of differential raw material use and miniaturization. However, similar methodologies were applied to studies using formal models for raw material and mobility studies applied to archaeology, where only artifacts above 2 cm were considered to improve comparative studies between assemblages [@surovell_toward_2009].

The sample from the Terrace area (VB-T) consists of 3600 lithic artifacts from levels 4 to 7/8 (excluding levels with a small sample such as level 4B), while the sample from the Shelter area (VBS) consists of ++ lithic artifacts from levels C to A. These two areas were chosen since both have detailed spatial information (as mentioned in section *IN*), several radiocarbon dates through different levels, and previous studies providing lithic technology, faunal, and geoarchaeological data. The archaeological materials are currently located at ICArEHB's laboratories (University of Algarve) and no permits were necessary for their study. The macroscopic analysis was divided into two steps. The first step included a preliminary characterization using a hand lens of 10x magnification. This step allowed us to create groups and subgroups based on similar macroscopic characteristics which have been used in previous studies to describe and discern between different types of chert; these include characteristics such as color, translucency, and feel [@luedtke_archaeologists_1992]; a data dictionary for the recorded variables can be found in the Supplementary Materials (++). In the second step, we used a Nikon SMZ25 stereomicroscope to observe each sample in further detail, which allowed us to better characterize the artifacts (especially regarding inclusions and alterations), and the previously established groups. The data dictionary used for the individual samples can also be found in Supplementary Materials (\*\*). Given the inherent macroscopic variability of the geological samples and archaeological specimens, as well as the number of samples, the database used for the individual analysis is simplified and focused on collecting data related to weight, cortex, and alterations. A caveat of using this methodology of analysis is the inability to unequivocally characterize and group artifacts which have surface alterations such as patinas, pits or fire-related alterations. Whenever artifacts displayed extensive alterations and could not be unequivocally characterized (@fig-terrace-alterations), they were analysed using the individual sample dataset, and instead grouped in a category of indeterminate cherts (INDET).

```{r}
#| label: fig-terrace-alterations
#| fig-cap: Photo and close-up of samples with extensive alterations which impossibilitate the identification of chert type. a) artifact with cracks, white patina and reddening covering 100% of the surface; b) artifact with surface alterations, displaying a portion with patina (brown) and a portion without patina (gray), cleaned with HCL.

knitr::include_graphics(here::here("analysis/figures/terrace_alterations.png"))

@@

To better characterize the different chert groups identified in the macroscopic analysis, 22 thin sections were prepared at Thin Section Lab (Toul, France) and analyzed at the ICArEHB laboratory. The samples chosen for this analysis include at least one sample from each group present throughout the stratigraphy of the Terrace area, from different archaeological levels (@tbl-thinsections). For groups that showed macroscopic variability, several thin sections were analyzed to ensure the collection of petrographic data that reflected this variability. Petrographic thin sections were analyzed by polarized light microscopy (Nikon LV100ND) to determine the mineral composition, textural characteristics, allochems, and bioclasts, as well as any alterations.

| Type               | Sample ID | Level | TS \# |
|--------------------|-----------|-------|-------|
| Chalcedony         | I19-1615  | 4D    | 8     |
| Type 2             | I19-2835  | 5     | 17    |
| Type 2 (rmu_17)    | I19-2226  | 4E    | 18    |
| Type 2 (altered)   | H18-1938  | 4D    | 20    |
| Type 2 (reddening) | H19-2426  | 4D    | 3     |
| Type 2B            | H21-3095  | 4C    | 6     |
| Type 2B            | H21-4234  | 5     | 7     |
| Type 15            | I20-3160  | 5     | 10    |
| Type 7             | H20-2441  | 4E    | 19    |
| Type 12            | I21-3252  | 4E    | 14    |
| Type 12B           | H18-2708  | 5     | 2     |
| CPT                | I19-3350  | 6     | 9     |
| CPT                | J18-778   | 5     | 13    |
| CPT                | I20-3689  | 6     | 16    |
| Sub-type LLB       | I20-3951  | 6     | 11    |
| Sub-type 11E       | H19-4074  | 6     | 4     |
| Type 11F           | H19-2924  | 4E    | 1     |
| Type 13 (01b; 01a) | L19-64    | 5     | 12    |
| Type 13 (04)       | J18-1264  | 8     | 22    |
| Type 13B           | I21-2966  | 4C    | 15    |
| Oolitic            | H19-4216  | 6     | 5     |
| White patinated    | H20-4166  | 6     | 21    |

: A table with thin sections {#tbl-thinsections}

The petrographic analysis PDF files for all thin sections studied in this paper (including macroscopic and microscopic figures) can be found in our GitHub page (++) as well as the OSF repository (++). The geological thin sections can also be consulted in the LusoLit online database at https://lusolit.icarehb.com/. A detailed macroscopic and petrographic description and figures for each type and sub-type of chert and chalcedony can be found in the SOM (++), as well as the datadictionary used for the petrographic analysis. The entirety of the R code used for the analysis and visual representations contained in this paper can be accessed through our online research compendium (++). We use the rrtools package by Ben Marwick to create a research compendium and write a reproducible journal article; rrtools is a free and open-source RStudio package accessible at <https://github.com/benmarwick/rrtools>. To produce these files, we followed the procedures outlined by @Marwick2017 for constructing research compendiums, aiming to enhance the reproducibility of our research. The provided files include the complete set of raw data used in the analysis, along with a custom R project [@wickhamPackagesOrganizeTest2015] containing the code required to generate all tables and figures. To enable maximum reuse, the code is made available under the MIT license, data under CC‐0, and figures under CC‐BY [additional details can be found in @Marwick2017].

@@

T_dataset <- T_dataset %>% 
  #filter(Level %in% c("4", "4C", "4D", "4E", "5", "6", "7"))
  filter(Level !="4B") %>%  # Filtering out levels with barely any artifacts
  mutate(Level = case_when(Level == "7C" ~ "7", # Collapsing samples from Level 7C into 7
                           TRUE ~ Level)) %>% 
  mutate(Level = case_when(Level == 8 ~ "7", # Collapsing two samples from Level 8 into 7, since 8 disappears in field_data
                           TRUE ~ Level))

#cols_to_concat_spit <- c("Level", "Spit")
#RM_by_spit <- T_dataset %>%
  #filter(!Spit %in% c("limp_corte", "profile", " ", NA)) %>%
  #select(ID, Z, Level, Spit, RMTYPE) %>%
  #unite(col='depth', cols_to_concat_spit, sep=".", remove=FALSE) %>% 
  #dplyr::group_by(depth) %>%
  #dplyr::count(RMTYPE) %>%
  #mutate(freq = n / sum(n)*100)

@@

# Shelter DB preparation
S_RM_data <- read_csv(here::here('analysis/data/raw_data/MACRO_IND_SHELTER.csv')) # Shelter RM analysis
S_field_data <- read_csv(here::here('analysis/data/raw_data/VB_Shelter_xyz_codes.csv')) # Shelter coordinates
S_layerdata <- read_csv(here::here('analysis/data/raw_data/shelter_layers_JC.csv'))

@@

S_layerdata <- S_layerdata %>% 
  mutate_at("Date", str_replace, "20", "") %>%
  dplyr::rename("YEAR" = "Date") %>% 
  unite(col='ID', cols_to_concat_shelter, sep="-", remove=FALSE) %>% 
  select(ID, Layer)

S_field_data <- S_field_data %>%
  mutate_at("YEAR", str_replace, "20", "") %>%
  unite(col='ID', cols_to_concat_shelter, sep="-", remove=FALSE) %>% # Unite YEAR with Id s«using cols_to_concat
  select(ID, X, Y, Z, CODE) %>% # Selecting important variables
  distinct(ID, .keep_all = TRUE)

shelter_field_data <- full_join(S_field_data, S_layerdata, by = "ID")
S_dataset <- left_join(S_RM_data, shelter_field_data, by = "ID", relationship = "many-to-many")

@@

The three main used raw materials at the archaeological site of Vale Boi are quartz, greywacke, and chert. Other raw materials such as chalcedony, dolerite and schists have also been identified. Using the field database with all individually coordinated lithic artifacts, we compared the incidence of chert and chalcedony in opposition to the remaining lithics of other raw materials. @fig-rm-comparison shows that for the selected levels, chert represents only \~20-15% of the lithic assemblage. Based on previous lithic studies in the terrace area, the remaining 80% of other raw materials are mostly composed of quartz and greywacke \[++\]. Although no systematic raw material research has been applied to the latter raw materials, they have been interpreted as local and readily available in the proximity of the site \[++\]. Frequently, the quartz and greywacke assemblages include large amounts of shattered and/or unknapped chunks or slabs. Previous studies from the terrace area have also shown that when removing shatter and chunks, the percentages of chert become much more representative when compared to greywacke and quartz [@belmiro_gravettian-solutrean_2021]. Based on this, despite the low amounts of individually plotted chert artifacts in the field database, chert continues to be one of the main raw materials to produce lithic stone tools, with complete knapping sequences and formal toolkits. The importance of chalcedony has also been shown in its use to produce formal tool kits during specific technocomplexes at the site [@belmiro_gravettian-solutrean_2021].

@@

@@

As previously mentioned, whenever too altered to confidently attribute a group, the artifacts were instead classified as indeterminable (INDET). As seen in @tbl-indet-freq, level 4 shows the higher percentages of indeterminate artifacts, corresponding to 40% of the total sample (n=1027). This is due to the frequent alterations of the sample, especially related to patinas and alteration rinds, pits and fire/heat alterations. For all other levels in the Terrace, indeterminate represents in mean \~28% of the chert artifacts, while in the Shelter the frequency of indetermiates is slightly lower (\~20-25%).

terrace_indet_table <- phase_dataset %>% 
  mutate(Group = if_else(RMTYPE=="INDET", "Indeterminate", "Grouped")) %>% 
  tabyl(Group,Level) %>% 
  adorn_totals("row") %>%
  adorn_percentages("col") %>%
  adorn_pct_formatting(2) %>%
  adorn_ns("front")

shelter_indet_table <- S_dataset %>% 
  mutate(Group = if_else(RMTYPE=="INDET", "Indeterminate", "Grouped")) %>% 
  mutate_at(vars(Layer), ~replace_na(., "ND")) %>% 
  filter(!Layer %in% c("D", "ND")) %>% 
  tabyl(Group,Layer) %>% 
  adorn_totals("row") %>%
  adorn_percentages("col") %>%
  adorn_pct_formatting(2) %>%
  adorn_ns("front")

indet_table <- full_join(terrace_indet_table, shelter_indet_table, by = "Group")

kable(head(indet_table))

## Characterization and source attribution

Through the macroscopic and petrographic analysis, we identified 10 main chert/chalcedony groups, present throughout the stratigraphy in the Terrace and Shelter area. A single different group was identified in the Shelter area, with limited presence in the Terrace (n=2), and 14 varieties of chert and 2 chalcedony composed of one to three samples were identified in several levels of the Terrace and Shelter. The latter varieties were collapsed into a category named TL (Trace lithotypes). Through the comparison with the reference collections, we identified cherts congruent with the local sources and cherts which are not local.

The raw materials which are congruent with local sources are the chalcedony and massive, micro-cryptocrystalline cherts. The chalcedony is a massive, fibrous chalcedony without allochems. It is characterized by a colorless and translucent appearance, often partially or completely white patinated. It is characterized by frequent large crystallizations, irregularities and fractures which impact the knapping. It is macroscopically and petrographically congruent with the hydrothermal chalcedony found in lenses in Triassic formations, also frequently characterized by irregularities and fractures.

The cherts are massive micro-cryptocrystalline quartzes from a marine environment. Macroscopically, these groups show variability which resulted in their separation in four different macroscopic groups (Group 2 to 5). In general, these are opaque to sub-translucent cherts, with colors varying brown, yellow, pale red and grey, with the frequent presence of spicules and iron oxides. The differences are in their colors, nodule morphology, frequency, and variety of bioclasts, something already identified in the geological samples [@belmiro2023]. Petrographically, however, these cherts are homogeneous, except for a small number of artifacts with a banded variety within group 5. The macroscopic separation was maintained for its possible usefulness in identifying preferred types of nodules or macroscopic varieties through time and for future correlation with technological data. These cherts are characterized by wackestone textures, with accessory amounts of macrocrystalline quartz and fibrous chalcedony, frequently found replacing fossils and occasionally dolomite. Group 12 shows higher percentages of micrite/sparite. In all samples, fossils range from common to very frequent, albeit frequently unidentifiable. When identifiable in thin-section, they are sponge spicules, with the rare occurrence of possible foraminifera and ostracod in a sample from type 12. Porosity varies between 10% and 1%, of vug type. Based on the previously described macroscopic and petrographic characteristics of the Lower Jurassic cherts of the Algarve [@belmiro2023], we interpret Groups 2, 15, 7 and 12 to be local, as they are congruent with the described characteristics. Groups 2 and 15 are congruent with the frequently available brown/yellow and red nodules and beds, ranging from uncommon to very frequent fossils. The macroscopic similarities of Groups 7 and 12 are more limited, as they show similarities to specific nodules, rarely found in the known outcrops, and mostly located in the Ferrel area. As such, we interpret these cherts as local, but from nodules and outcrops that are no longer frequently available or visible in the landscape. This is especially relevant, since the Ferrel outcrops are mostly destroyed and can be found in poor preservation conditions, and with small, broken nodules in the surface, frequently altered [@belmiro2023].

Five chert types, not congruent with local sources, were identified (Groups 6 to 10), including the TL group.
